---
title: "Ocular_snRNAseq_li_etal_2024_mouse_retina"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: yes
    code_folding: hide
---

# Abstract
This dataset was analyzed for the purpose of comparing GRK1 expression in the major two subtypes of photoreceptors (Rods vs Cones).   

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                    warning = FALSE,
                    comment = "")
```

```{r}
library(biomaRt)
library(Seurat)
library(SeuratDisk)
library(BPCells)
library(Azimuth)
library(anndata)
library(scCustomize)
library(sctransform)
library(glmGamPoi)
library(tidyverse)
library(reshape2)
library(corrplot)
library(future)
options(future.globals.maxSize = 800 * 1024^2) # tested, use 600M
library(openxlsx)
library(patchwork)
library(HGNChelper)
library(httpgd)
library(Scillus)
library(fgsea)
library(msigdbr)
library(qs)
library(knitr)
Sys.setenv(PROJ_LIB = "/home/ubuntu/anaconda3/envs/r4.2.1/share/proj")
Sys.setenv(GDAL_DATA = "/home/ubuntu/anaconda3/envs/r4.2.1/share/gdal")
source("/mnt/data/apps/sc-type/R/gene_sets_prepare.R")
source("/mnt/data/apps/sc-type/R/sctype_score_.R")
setwd('~/Ocular_snRNAseq_publicdata_GRK1/mouse/li_etal_2024_mouse_retina')
set.seed(123)
```


# Load downloaded count matrix object from cellxgene  
```{r, eval = FALSE}
# retina.data <- open_matrix_anndata_hdf5('li_etal_2023_human_retina.h5ad')
# write_matrix_dir(mat = retina.data, dir = 'li_etal_2023_human_retina_counts')
# retina.mat <- open_matrix_dir(dir = "li_etal_2023_human_retina_counts")

# retina.mat <- Azimuth:::ConvertEnsembleToSymbol(mat = retina.mat, species = "human")

# m <- read.csv("li_etal_2023_human_retina.metadata.csv", header=TRUE)
# retina <- CreateSeuratObject(counts = retina.mat, meta.data = m)

retina <- readRDS('li_etal_2024_mouse_retina.rds')
retina.mat <- retina[['RNA']]$counts

retina.mat <- Azimuth:::ConvertEnsembleToSymbol(mat = retina.mat, species = "mouse")

retina <- CreateSeuratObject(counts = retina.mat, meta.data = retina[[]])

retina@meta.data <- retina@meta.data %>%
    rename('sample' = 'donor_id')
```

```{r}
retina <- qread('retina.qs')
```

# QC and selecting cells for further analysis
## QC

- The object after pre-process
```{r}
retina
table(retina$sample)
table(retina$majorclass)
table(retina$cell_type)
head(retina, 5)
```

## QC stats
```{r}
summary(retina$nCount_RNA)
summary(retina$nFeature_RNA)
summary(retina$percent.mt)
```

** select samples that have photoreceptors**
```{r, eval = FALSE}
# they already set percent.mt<10, if lower the cutoff many Cones will be filtered out
retina <- subset(retina, subset = percent.mt < 10 &
    nFeature_RNA > 100 &
    nCount_RNA > 200)

samples <- unique(retina$sample)
selected <- samples[grepl("WT", samples)]
retina <- subset(retina, subset = sample %in% selected)
```

## QC metrics as a violin plot 
- by donor
```{r, fig.height=6, fig.width=12}
VlnPlot(retina, layer = "counts", features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
    group.by = 'sample',
    pt.size = 0, ncol = 3)
```


# SCTransform normalization, PCA and Clustering 
```{r, eval = FALSE}
# regressing out unwanted variation (mitochondrial, cell cyle, etc.)
retina <- SCTransform(retina, method = "glmGamPoi", vst.flavor = "v2",
        vars.to.regress = c("percent.mt"), verbose = TRUE, return.only.var.genes = FALSE)

retina <- RunPCA(retina, npcs = 40)
# ElbowPlot(retina, ndims = 40)
retina <- RunUMAP(retina, reduction = "pca", dims = 1:30)
retina <- FindNeighbors(retina, reduction = "pca", dims = 1:30)
retina <- FindClusters(retina, resolution = 0.1)
table(retina$seurat_clusters)
```

```{r, eval = FALSE}
qsave(retina, file = 'retina.qs')
```


# Cell type annotation
## Apply marker based tool `sc-type`
```{r}
sctype <- function(obj) {
    # prepare marker genes from the database, category: Immune system, Liver, Pancreas, Kidney, Eye, Brain
    gs_list <- gene_sets_prepare("/mnt/data/apps/sc-type/ScTypeDB_short_NHPeye.xlsx", "Eye")

    # use a customized marker list
    # gs_list <- gene_sets_prepare("./ocular_customized_markers.xlsx", "Eye")
    # get cell-type by cell matrix
    # NOTE: scRNAseqData parameter should correspond to your input scRNA-seq matrix.
    # In case Seurat is used, it is either pbmc[["RNA"]]@scale.data (default),
    # pbmc[["SCT"]]@scale.data, in case sctransform is used for normalization,
    # or pbmc[["integrated"]]@scale.data, in case a joint analysis of multiple single-cell datasets is performed.
    es.max <- sctype_score(scRNAseqData = obj[['SCT']]@scale.data, scaled = TRUE,
                        gs = gs_list$gs_positive, gs2 = gs_list$gs_negative)

    # merge by cluster
    metadata <- obj@meta.data
    clusters <- unique(obj@meta.data$seurat_clusters)
    cL_resutls <- do.call("rbind",
        lapply(
            clusters,
            function(cl) {
                es.max.cl <- sort(rowSums(es.max[, rownames(metadata[metadata$seurat_clusters == cl, ])]),
                    decreasing = TRUE)
                cell_scores <- data.frame(cluster = cl,
                        type = names(es.max.cl),
                        scores = es.max.cl,
                        ncells = sum(metadata$seurat_clusters == cl))
                head(cell_scores, 10)
            }
        )
    )
    sctype_scores <- cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)

    # set low-confident (low ScType score) clusters to "unknown"
    sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] <- "Unknown"
    print(data.frame(sctype_scores))

    # add cell type annotations to obj
    obj@meta.data$celltypes <- ""
    for (j in unique(sctype_scores$cluster)){
    cl_type <- sctype_scores[sctype_scores$cluster == j, ]
    obj@meta.data$celltypes[obj@meta.data$seurat_clusters == j] <-
        as.character(cl_type$type[1])
    }
    return(obj)
}
```

```{r, eval = FALSE}
retina <- sctype(retina)
```


## Show cell types on UMAP
```{r, fig.height=12, fig.width=12}
DimPlot(retina, reduction = "umap", label = TRUE, repel = TRUE,
    group.by = 'cell_type') +
    theme(legend.position = "top")
table(retina$cell_type)

DimPlot(retina, reduction = "umap", label = TRUE, repel = TRUE,
    group.by = 'majorclass') +
    theme(legend.position = "top")
table(retina$majorclass)
```


## Number of cells in each cell type
```{r}
plot_celltypes <- function(srat, celltype_col) {
    cell_type <- srat@meta.data[[celltype_col]]
    count_table <- table(srat@meta.data[[celltype_col]], srat@meta.data$sample)
    count_mtx <- as.data.frame.matrix(count_table)
    count_mtx$cell_type <- rownames(count_mtx)
    melt_mtx <- melt(count_mtx)

    celltype_size <- aggregate(value ~ cell_type, data = melt_mtx, FUN = sum)
    celltype_size <- celltype_size %>% dplyr::arrange(value)
    celltype_size$cell_type <- factor(celltype_size$cell_type, levels = celltype_size$cell_type)
    melt_mtx$cell_type <- factor(melt_mtx$cell_type, levels = celltype_size$cell_type)
    colnames(melt_mtx)[2] <- "sample"

    p1 <- ggplot(celltype_size, aes(y = cell_type, x = value)) +
    geom_bar(position = "dodge", stat = "identity", fill = "grey60") +
    theme_bw() + scale_x_log10() + xlab("Cells per celltype, log10 scale") + ylab("")

    p2 <- ggplot(melt_mtx, aes(x = cell_type, y = value, fill = sample)) +
        geom_bar(position = "fill", stat = "identity") + theme_bw() + coord_flip() +
        scale_fill_brewer(palette = "Set2") +
        ylab("Fraction of cells in each sample") + xlab("celltype number") + theme(legend.position = "top")

    p2 + p1 + plot_layout(widths = c(3, 1))
}
```

```{r, fig.height=6, fig.width=14}
plot_celltypes(retina, 'majorclass')
```

## plot % Rods and Cones for each sample/tissue/source
```{r, fig.width=6, fig.height=6}
Idents(retina) <- "majorclass"
cell_type_to_plot <- c("Rod", 'Cone')
percentage_data <-retina@meta.data %>%
    count(sample, majorclass) %>%
    # Calculate the percentage
    group_by(sample) %>%
    mutate(Percentage = n / sum(n)) %>%
    filter(majorclass %in% cell_type_to_plot) %>%
    ungroup()

# pdf("Percent_RGC_per_sample.pdf", 8,6)
ggplot(percentage_data, aes(x = sample, y = Percentage, fill = sample)) +
    geom_bar(stat = "identity") +
    labs(title = paste("Percentage of cells"), fill = "sample") +
    theme_minimal() +
    facet_wrap(~majorclass) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1), plot.title = element_text(hjust = 0.5)) +
    scale_y_continuous(labels = scales::percent) +
    theme(plot.margin = margin(5, 5, 5, 20, "mm")) 
# dev.off()
```

#  Cell markers analysis
== Load all cluster markers obj
```{r}
all.markers <- readRDS("all.cluster_markers.rds")
de_genes <- readRDS("de_genes_rod_vs_cone.rds")
```

## Find all markers
```{r, eval = FALSE}
Idents(retina) <- 'majorclass'
retina <- PrepSCTFindMarkers(retina, assay = "SCT", verbose = TRUE)
plan("multicore", workers = 12)
all.markers <- FindAllMarkers(retina, assay = "SCT", min.pct = 0.1, verbose = TRUE)
plan(sequential)
```

### Save marker genes
```{r, eval = FALSE}
all.markers <- all.markers %>% relocate(gene)
write_tsv(all.markers, "all.cluster_markers.tsv")
saveRDS(all.markers, file = "all.cluster_markers.rds")
```

### list top50 markers for Rods and Cones
**Rods**  
```{r}
top50 <- all.markers %>%
    group_by(cluster) %>%
    slice_max(order_by = avg_log2FC, n = 50)
    
kable(top50 %>% filter(cluster %in% 'Rod'))
```

**Cones**  
```{r}
kable(top50 %>% filter(cluster %in% 'Cone'))
```

**GRK1 stats**
# ```{r}
# kable(all.markers[all.markers$gene == "Grk1" & all.markers$cluster == 'Rod', ])
# ```

## Check Photoreceptors markers
### Violin plots
```{r, fig.height=14, fig.width=14}
rod_markers <- c("Esrrb","Cnga1", "Rho", "Abca4")
cone_markers <- c("Pde6c", "Kcne2", "Gnat2", "Arr3")
VlnPlot(object = retina, features = c("Grk1", rod_markers, cone_markers), pt.size = 0.01, combine = TRUE)
```

### UMAPs
```{r, fig.height=8, fig.width=8}
DimPlot(retina, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'majorclass') + theme(legend.position = "top")
```

```{r, fig.height=14, fig.width=14}
FeaturePlot(object = retina, features = c("Grk1", rod_markers, cone_markers), label =TRUE, combine = TRUE)
```

### Dot plot for markers in each cell type
- showing top markers with customized markers
```{r, fig.height=8, fig.width=10}
Idents(retina) <- "majorclass"
DefaultAssay(retina) <- "SCT"
top_genes <- all.markers %>%
    filter(cluster %in% c('Rod', 'Cone')) %>%
    group_by(cluster) %>%
    slice_max(order_by = avg_log2FC, n = 10)

# Specify any additional genes you want to label
additional_genes <- c("Grk1", "Rho", "Abca4", "Arr3")

# Combine top genes and additional genes, ensuring uniqueness
label_genes <- unique(c(top_genes$gene, additional_genes))

DotPlot(retina, features = label_genes, cols = c("white", 'blue'), dot.scale = 8) +
    RotatedAxis()
```


# Differential expression between Rods and Cones
```{r, eval = FALSE}
plan("multicore", workers = 12)
de_genes <- FindMarkers(retina, ident.1 = 'Rod', ident.2 = 'Cone', min.pct = 0.01, assay = "SCT", verbose = TRUE)
plan(sequential)

de_genes <- rownames_to_column(de_genes, var = "gene")
write_tsv(de_genes, "de_genes_rod_vs_cone.tsv")
saveRDS(de_genes, file = "de_genes_rod_vs_cone.rds")
```

```{r}
up_gene <- de_genes %>%
    slice_max(order_by = avg_log2FC, n = 50)

down_gene <- de_genes %>%
    slice_min(order_by = avg_log2FC, n = 50)
```

**Significantly up-regulated genes in Rods(top50)**
```{r}
kable(up_gene)
```
**Significantly up-regulated genes in Cones(top50)**
```{r}
kable(down_gene)
```

**GRK1 stats**
```{r}
kable(de_genes[de_genes$gene == "Grk1", ])
```

## volcano plot
```{r}
# Add a column for -log10(p-value)
de_genes$p_val_adj_log10 <- -log10(de_genes$p_val_adj)

# Determine the top 10 genes based on adjusted P-value
top_genes <- de_genes %>%
    slice_max(order_by = avg_log2FC, n = 5)

bottom_genes <- de_genes %>%
    slice_min(order_by = avg_log2FC, n = 5)

# Specify any additional genes you want to label
additional_genes <- c("Grk1", "Rho", "Abca4", "Arr3")

# Combine top genes and additional genes, ensuring uniqueness
label_genes <- unique(c(top_genes$gene, bottom_genes$gene, additional_genes))

# Create the volcano plot
volcano_plot <- ggplot(de_genes, aes(x = avg_log2FC, y = p_val_adj_log10, label = gene)) +
    geom_point(alpha = 0.5) + 
    labs(title = "Volcano Plot of Differential Expression (Rod vs Cone)",
        x = "Average Log2 Fold Change",
        y = "-Log10 Adjusted P-Value") +
    geom_text(data = filter(de_genes, gene %in% label_genes), vjust = 1.5, hjust = 0.5,  color = "red", check_overlap = TRUE) +
    theme_minimal() +
    theme(plot.title = element_text(size = 16, hjust = 0.5, margin = margin(b = 20)))

# Print the plot
print(volcano_plot)

```


# Plots to report

## rename cell types
```{r}
rename_map <- c("MG" = "MC")
retina$majorclass <- factor(retina$majorclass)
levels(retina$majorclass) <- ifelse(levels(retina$majorclass) %in% names(rename_map), 
                                    rename_map[levels(retina$majorclass)], 
                                    levels(retina$majorclass))

Idents(retina) <- "majorclass"
cell_type <- c("Rod", "Cone", "HC", "BC", "AC", "RGC", "MC", "RPE")
sub_retina <- subset(retina, idents = cell_type)
table(sub_retina$majorclass)
```

## DotPlot for selected markers
```{r}
DefaultAssay(sub_retina) <- "SCT"
Idents(sub_retina) <- "majorclass"
sub_retina$majorclass <- factor(sub_retina$majorclass, levels = cell_type)

markers <- c("Grk1", "Abca4", "Rho", "Nrl", "Gnat1", "Arr3", "Gnat2", 
    "Opn1sw", "Opn1mw", "Onecut1", "Vsx2", "Gad1", "Rbpms", "Rlbp1", "Rpe65", "Best1")

DotPlot(sub_retina, features = rev(markers), cols = c("white", 'blue'), group.by = "majorclass", dot.scale = 10, scale = TRUE) + RotatedAxis() + coord_flip()
```

## Heatmap for selected markers
** randomly sample 100 cells for each cell type **
```{r}
DoHeatmap(subset(sub_retina, downsample = 100), features = markers, group.by = "majorclass", size = 3, slot = "scale.data") + 
    theme(plot.margin = margin(t = 20))

```

** sample 10% of cells for each cell type **
```{r}
sampled_sub_seurat <- qread('sampled_sub_seurat.qs')
table(sampled_sub_seurat$majorclass)
```

```{r, eval = FALSE}
seurat_list <- SplitObject(sub_retina, split.by = "majorclass")
# Sample 10% of cells from each cell type
sampled_list <- lapply(seurat_list, function(x) {
    # Calculate number of cells to sample
    num_to_sample <- ceiling(ncol(x) * 0.1)
    # Randomly sample cells
    sampled_cells <- sample(colnames(x), num_to_sample)
    # Subset the Seurat object for sampled cells
    subset(x, cells = sampled_cells)
})
# Merge the sampled objects back into a single Seurat object
sampled_sub_seurat <- Reduce(function(x, y) merge(x, y), sampled_list)
table(sampled_sub_seurat$majorclass)
```

```{r, eval = FALSE}
qsave(sampled_sub_seurat, file = 'sampled_sub_seurat.qs')
```


```{r}
Idents(sampled_sub_seurat) <- "majorclass"
cell_type <- c("Rod", "Cone", "HC", "BC", "AC", "RGC", "MC", "RPE")
sampled_sub_seurat$majorclass <- factor(sampled_sub_seurat$majorclass, levels = cell_type)
levels(sampled_sub_seurat$majorclass)

DoHeatmap(sampled_sub_seurat, features = markers, group.by = "majorclass", size = 3,  slot = "scale.data") +
    theme(plot.margin = margin(t = 20))

```
