---
title: "Ocular snRNAseq pilot_dataset1+dataset2"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: yes
    code_folding: hide
---

# Abstract
This is our pilot Ocualr Single nucleus RNAseq for Glaucoma disease.
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      comment = "")
```

```{r}
library(biomaRt)
library(Seurat)
library(sctransform)
library(glmGamPoi)
library(tidyverse)
library(reshape2)
library(future)
library(openxlsx)
library(patchwork)
library(HGNChelper)
Sys.setenv(PROJ_LIB = "/home/ubuntu/anaconda3/envs/r4.2.1/share/proj")
Sys.setenv(GDAL_DATA = "/home/ubuntu/anaconda3/envs/r4.2.1/share/gdal")
source("/mnt/data/apps/sc-type/R/gene_sets_prepare.R")
source("/mnt/data/apps/sc-type/R/sctype_score_.R")
setwd('~/Ocular_snRNAseq_pilot_2nd')
set.seed(123)
```


# Load and combine sample feature_bc_matrix from cellranger 
```{r, eval = FALSE}
# List all the files with ".rds$" and "_S" in their names
rds_path1 <- list.files('../Ocular_snRNAseq_pilot/nf-core/output/cellranger/cellranger/mtx_conversions/',
    pattern = "NHP_S_[1-4]_matrix\\.rds", full.names = TRUE)
rds_names1 <-  gsub("_matrix.rds$", "", basename(rds_path1))
rds_list1 <- lapply(rds_path1, readRDS)
names(rds_list1) <- rds_names1

# dataset2
rds_path2 <- list.files('./nf-core/output/cellranger/cellranger/mtx_conversions/',
    pattern = "DAPI_[1-3]_matrix\\.rds", full.names = TRUE)
rds_names2 <-  gsub("_matrix.rds$", "", basename(rds_path2))
rds_list2 <- lapply(rds_path2, readRDS)
names(rds_list2) <- rds_names2

rds_list <- c(rds_list1, rds_list2)

# combine samples
seuratObj <- merge(rds_list[[1]], y = rds_list[2:7], add.cell.ids = names(rds_list), project = "ocular_pilot")
# add sample names
cells <- rownames(seuratObj@meta.data)

tmp <- sub("(NHP_S_[1-4]).*", "\\1", cells)
seuratObj$sample <- sub("(DAPI_[1-4]).*", "\\1", tmp)

#add groups
dataset_info <- ifelse(grepl("NHP", seuratObj$sample), "dataset1", ifelse(grepl("DAPI", seuratObj$sample), "dataset2", "unknown"))
seuratObj <- AddMetaData(seuratObj, metadata = dataset_info, col.name = "dataset")

```

# QC and selecting cells for further analysis

## Add %mitochondrial counts and cell cycle scores 
```{r, eval = FALSE}
## mitochondrial genes:
mt_genes <- c("ND1", "ND2", "COX1", "COX2", "ATP8", "ATP6", "COX3", "ND3", "ND4L", "ND4", "ND5", "ND6", "CYTB")

seuratObj[["percent.mt"]] <- PercentageFeatureSet(seuratObj, features = mt_genes)

s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

seuratObj <- CellCycleScoring(seuratObj, s.features = s.genes, g2m.features = g2m.genes)
```

## QC
- Filter cells by:  
1, umi/nCount_RNA, low-quality cells (very low UMI counts) or doublets (very high UMI counts)  
2, gene/nFeature_RNA, remove potential low-quality cells and multiplets as above  
3, percent.mt, note that this step directly removes cells with high mt, the regress.out function later   
removes variations from mt and cell-cyle gene expression.   
4, our cutoff is chosen by visulizing data in Loupe browser    

== Load filtered Seurat obj instead of re-run the following steps 
```{r}
seuratObj <- readRDS("ocular_obj.rds")
```

- The object after pre-process
```{r}
seuratObj
table(seuratObj[[]]$sample)
table(seuratObj[[]]$dataset)
head(seuratObj, 5)
```

### QC stats
```{r}
summary(seuratObj$nCount_RNA)
summary(seuratObj$nFeature_RNA)
summary(seuratObj$percent.mt)
```

UMI range is (`r min(seuratObj$nCount_RNA)`, `r max(seuratObj$nCount_RNA)`)
nfeatue range is (`r min(seuratObj$nFeature_RNA)`, `r max(seuratObj$nFeature_RNA)`)

Retinal ganglion cells are larger, results in a greater amount of total RNA per cell. 
Skip the nCount_RNA and nFeature_RNA filter as the gRed team did in their 2023 Cell Genomics paper.

Only remove cells with >5% mitochondrial expressions: `r length(seuratObj$percent.mt[seuratObj$percent.mt>5])`

```{r, eval = FALSE}
seuratObj <- subset(seuratObj, subset = percent.mt < 5)
```

### QC metrics as a violin plot 
- by initial clusters
```{r, fig.height=10, fig.width=12}
VlnPlot(seuratObj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), pt.size = 0, ncol = 1)
```

- by samples
```{r, fig.height=6, fig.width=12}
VlnPlot(seuratObj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
    group.by = 'sample',
    pt.size = 0.2, ncol = 3)
```


## SCTransform normalization, PCA and Clustering 
```{r, eval = FALSE}
# regressing out unwanted variation (mitochondrial, cell cyle, etc.)
seuratObj <- SCTransform(seuratObj, method = "glmGamPoi", vst.flavor = "v2",
        vars.to.regress = c("percent.mt", "S.Score", "G2M.Score"), verbose = TRUE)

seuratObj <- RunPCA(seuratObj, npcs = 30)
# ElbowPlot(seuratObj, ndims = 30)
seuratObj <- RunUMAP(seuratObj, reduction = "pca", dims = 1:30)
seuratObj <- FindNeighbors(seuratObj, reduction = "pca", dims = 1:30)
seuratObj <- FindClusters(seuratObj, resolution = 0.3)
```

== Save processed object
```{r, eval = FALSE}
saveRDS(seuratObj, file = "ocular_obj.rds")
```

## Visualize initial clusters
```{r, fig.height=7, fig.width=14}
p1 <- DimPlot(seuratObj, reduction = "umap", group.by = "sample") + theme(legend.position = "top")
p2 <- DimPlot(seuratObj, reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()
p1 + p2
```

```{r, fig.height=10, fig.width=14}
DimPlot(seuratObj, reduction = "umap", label = FALSE, split.by = "sample", ncol = 4) + theme(legend.position = "top")
```

## Number of cells in each cluster
```{r}
plot_clusters <- function(srat) {
    count_table <- table(srat@meta.data$seurat_clusters, srat@meta.data$sample)
    count_mtx <- as.data.frame.matrix(count_table)
    count_mtx$clusters <- rownames(count_mtx)
    melt_mtx <- melt(count_mtx)

    celltype_size <- aggregate(value ~ clusters, data = melt_mtx, FUN = sum)
    celltype_size <- celltype_size %>% dplyr::arrange(value)
    celltype_size$clusters <- factor(celltype_size$clusters, levels = celltype_size$clusters)
    melt_mtx$clusters <- factor(melt_mtx$clusters, levels = celltype_size$clusters)
    colnames(melt_mtx)[2] <- "sample"

    p1 <- ggplot(celltype_size, aes(y = clusters, x = value)) +
    geom_bar(position = "dodge", stat = "identity", fill = "grey60") +
    theme_bw() + scale_x_log10() + xlab("Cells per cluster, log10 scale") + ylab("")

    p2 <- ggplot(melt_mtx, aes(x = clusters, y = value, fill = sample)) +
        geom_bar(position = "fill", stat = "identity") + theme_bw() + coord_flip() +
        scale_fill_brewer(palette = "Set2") +
        ylab("Fraction of cells in each sample") + xlab("cluster number") + theme(legend.position = "top")

    p2 + p1 + plot_layout(widths = c(3, 1))
}
```
```{r fig.height=8, fig.width=14}
plot_clusters(seuratObj)
```

# Data Integration  
To identify cell types present in all samples/conditions/modalities within our dataset,  
and would like to observe a representation of cells from both samples/conditions/modalities in every cluster.  
Reference: https://github.com/hbctraining/scRNA-seq_online/blob/master/lessons/06_integration.md  
- Execute the normalization, variance estimation, and identification of the most variable genes for each sample   
- Perform integration of cells across conditions using the most variant genes to identify cells most similar to each other  

== Load object list `seuratObj.list`
```{r, eval = FALSE}
seuratObj.list <- readRDS("ocular_obj.list.rds")
```

== Load objects `ocular_obj.integrated`
```{r}
seuratObj.integrated <- readRDS("ocular_obj.integrated.rds")
table(seuratObj.integrated$seurat_clusters)
```

## Split object and SCTransform normalization  
split the dataset into a list of two seurat objects (GLA_KO and WT), 
normalize and identify variable features for each dataset independently.  
```{r, eval = FALSE}
seuratObj.list <- SplitObject(seuratObj, split.by = "sample")

for (i in 1:length(seuratObj.list)) {
    seuratObj.list[[i]] <- SCTransform(seuratObj.list[[i]], method = "glmGamPoi", vst.flavor = "v2",
        vars.to.regress = c("percent.mt", "S.Score", "G2M.Score"), verbose = TRUE) %>%
    RunPCA(npcs = 30)
}
```

## Select features and prepare integration 
Select features that are repeatedly variable across datasets for integration, 
and run `PrepSCTIntegration`, which ensures that all necessary Pearson residuals have been calculated.  
```{r, eval = FALSE}
seuratObj.features <- SelectIntegrationFeatures(object.list = seuratObj.list, nfeatures = 3000)
seuratObj.list <- PrepSCTIntegration(object.list = seuratObj.list, anchor.features = seuratObj.features)
```

== Save object `seuratObj.list`
```{r, eval = FALSE}
saveRDS(seuratObj.list, file = "ocular_obj.list.rds")
```

## Identify anchors and integrate the datasets
### using the 2nd dataset as references (samples 5-7) 
When using a set of specified references, anchors are first found between each query and each reference. 
The references are then integrated through pairwise integration. Each query is then mapped to the integrated reference.
Make sure to set normalization.method = **"SCT"**, and reduction = **"rpca"** for reciprocal PCA  
```{r, eval = FALSE}
anchors <- FindIntegrationAnchors(object.list = seuratObj.list, normalization.method = "SCT",
    reference = c(5, 6, 7),
    anchor.features = seuratObj.features, reduction = "rpca")
seuratObj.integrated <- IntegrateData(anchorset = anchors, normalization.method = "SCT")
```


# Cluster on the integrated dataset
## PCA and Clustering  
```{r, eval = FALSE}
DefaultAssay(seuratObj.integrated) <- "integrated"
seuratObj.integrated <- RunPCA(seuratObj.integrated, npcs = 30)
# ElbowPlot(seuratObj.integrated, ndims = 30)

seuratObj.integrated <- RunUMAP(seuratObj.integrated, reduction = "pca", dims = 1:30)
seuratObj.integrated <- FindNeighbors(seuratObj.integrated, reduction = "pca", dims = 1:30)
# try 0.3-0.5, give 26, 28, 30 clusters
seuratObj.integrated <- FindClusters(seuratObj.integrated, resolution = 0.3)
table(seuratObj.integrated$seurat_clusters)
```

== Save integrated object
```{r, eval = FALSE}
saveRDS(seuratObj.integrated, file = "ocular_obj.integrated.rds")
```

## Plots after integration   
```{r, fig.height=7, fig.width=14}
p1 <- DimPlot(seuratObj.integrated, reduction = "umap", group.by = "sample") + theme(legend.position = "top")
p2 <- DimPlot(seuratObj.integrated, reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()
p1 + p2
```

```{r, fig.height=10, fig.width=14}
DimPlot(seuratObj.integrated, reduction = "umap", label = FALSE, split.by = "sample", ncol = 4) + theme(legend.position = "top")
```

## Number of cells in each cluster
```{r}
plot_integrated_clusters <- function(srat) { 
    count_table <- table(srat@meta.data$seurat_clusters, srat@meta.data$sample)
    count_mtx <- as.data.frame.matrix(count_table)
    count_mtx$clusters <- rownames(count_mtx)
    melt_mtx <- melt(count_mtx)

    celltype_size <- aggregate(value ~ clusters, data = melt_mtx, FUN = sum)
    celltype_size <- celltype_size %>% dplyr::arrange(value)
    celltype_size$clusters <- factor(celltype_size$clusters, levels = celltype_size$clusters)
    melt_mtx$clusters <- factor(melt_mtx$clusters, levels = celltype_size$clusters)
    colnames(melt_mtx)[2] <- "sample"

    p1 <- ggplot(celltype_size, aes(y = clusters, x = value)) +
    geom_bar(position = "dodge", stat = "identity", fill = "grey60") +
    theme_bw() + scale_x_log10() + xlab("Cells per cluster, log10 scale") + ylab("")

    p2 <- ggplot(melt_mtx, aes(x = clusters, y = value, fill = sample)) +
        geom_bar(position = "fill", stat = "identity") + theme_bw() + coord_flip() +
        scale_fill_brewer(palette = "Set2") +
        ylab("Fraction of cells in each sample") + xlab("cluster number") + theme(legend.position = "top")

    p2 + p1 + plot_layout(widths = c(3, 1))
}
```
```{r fig.height=8, fig.width=14}
plot_integrated_clusters(seuratObj.integrated)
```


# Cell type annotation

## Apply marker based tool `sc-type`
```{r}
# prepare marker genes from the database, category: Immune system, Liver, Pancreas, Kidney, Eye, Brain
# gs_list <- gene_sets_prepare("/mnt/data/apps/sc-type/ScTypeDB_short.xlsx", "Eye")

# use a customized marker list
gs_list <- gene_sets_prepare("ocular_customized_markers.xlsx", "Eye")
# get cell-type by cell matrix
# NOTE: scRNAseqData parameter should correspond to your input scRNA-seq matrix.
# In case Seurat is used, it is either pbmc[["RNA"]]@scale.data (default),
# pbmc[["SCT"]]@scale.data, in case sctransform is used for normalization,
# or pbmc[["integrated"]]@scale.data, in case a joint analysis of multiple single-cell datasets is performed.
es.max <- sctype_score(scRNAseqData = seuratObj.integrated[['integrated']]@scale.data, scaled = TRUE,
                      gs = gs_list$gs_positive, gs2 = gs_list$gs_negative)

# merge by cluster
metadata <- seuratObj.integrated@meta.data
clusters <- unique(seuratObj.integrated@meta.data$seurat_clusters)
cL_resutls <- do.call("rbind",
    lapply(
        clusters,
        function(cl) {
            es.max.cl <- sort(rowSums(es.max[, rownames(metadata[metadata$seurat_clusters == cl, ])]),
                decreasing = TRUE)
            cell_scores <- data.frame(cluster = cl,
                    type = names(es.max.cl),
                    scores = es.max.cl,
                    ncells = sum(metadata$seurat_clusters == cl))
            head(cell_scores, 10)
        }
    )
)
sctype_scores <- cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)

# set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] <- "Unknown"
data.frame(sctype_scores)

# add cell type annotations to obj
seuratObj.integrated@meta.data$celltypes <- ""
for (j in unique(sctype_scores$cluster)){
  cl_type <- sctype_scores[sctype_scores$cluster == j, ]
  seuratObj.integrated@meta.data$celltypes[seuratObj.integrated@meta.data$seurat_clusters == j] <-
    as.character(cl_type$type[1])
}
```


## Show cell types on UMAP
```{r, fig.height=14, fig.width=14}
DimPlot(seuratObj.integrated, reduction = "umap", label = TRUE, repel = TRUE,
    group.by = 'celltypes') +
    theme(legend.position = "top")
```

```{r, fig.height=14, fig.width=14}
DimPlot(seuratObj.integrated, reduction = "umap", label = TRUE, repel = TRUE, split.by = 'sample', ncol = 4,
    group.by = 'celltypes') +
    theme(legend.position = "top")
```

## Number of cells in each cell type
```{r}
plot_celltypes <- function(srat) {
    count_table <- table(srat@meta.data$celltypes, srat@meta.data$sample)
    count_mtx <- as.data.frame.matrix(count_table)
    count_mtx$celltypes <- rownames(count_mtx)
    melt_mtx <- melt(count_mtx)

    celltype_size <- aggregate(value ~ celltypes, data = melt_mtx, FUN = sum)
    celltype_size <- celltype_size %>% dplyr::arrange(value)
    celltype_size$celltypes <- factor(celltype_size$celltypes, levels = celltype_size$celltypes)
    melt_mtx$celltypes <- factor(melt_mtx$celltypes, levels = celltype_size$celltypes)
    colnames(melt_mtx)[2] <- "sample"

    p1 <- ggplot(celltype_size, aes(y = celltypes, x = value)) +
    geom_bar(position = "dodge", stat = "identity", fill = "grey60") +
    theme_bw() + scale_x_log10() + xlab("Cells per celltype, log10 scale") + ylab("")

    p2 <- ggplot(melt_mtx, aes(x = celltypes, y = value, fill = sample)) +
        geom_bar(position = "fill", stat = "identity") + theme_bw() + coord_flip() +
        scale_fill_brewer(palette = "Set2") +
        ylab("Fraction of cells in each sample") + xlab("celltype number") + theme(legend.position = "top")

    p2 + p1 + plot_layout(widths = c(3, 1))
}
```

```{r fig.height=6, fig.width=14}
plot_celltypes(seuratObj.integrated)
```



# Cluster markers analysis
== Load all cluster markers obj
```{r}
all.markers <- readRDS("all.cluster_markers.rds")
head(all.markers)
```

## Find all markers
```{r, eval = FALSE}
seuratObj.integrated <- PrepSCTFindMarkers(seuratObj.integrated, assay = "SCT", verbose = TRUE)
# plan(multicore)
all.markers <- FindAllMarkers(seuratObj.integrated, assay = "SCT", verbose = TRUE)
# plan(sequential)
```

## Save marker genes
```{r, eval = FALSE}
all.markers <- all.markers %>% relocate(gene)
write_tsv(all.markers, "all.cluster_markers.tsv")
top20 <- all.markers %>%
    group_by(cluster) %>%
    top_n(n = 25, wt = avg_log2FC)
write_tsv(top20, "top25.cluster_markers.tsv")
saveRDS(all.markers, file = "all.cluster_markers.rds")
```

## Heatmap with top genes {.tabset}

### Markers for each Cluster
- showing expression in 10% random cells for each cluster  
```{r, fig.height=20, fig.width=12}
Idents(seuratObj.integrated) <- "seurat_clusters"

top10 <- all.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC)

DoHeatmap(subset(seuratObj.integrated, downsample = 100), features = top10$gene, size = 3) +
    theme(legend.position = 'none', plot.margin = margin(t = 20))
```