---
title: "Ocular_snRNAseq_Aflibercept_NHP_1"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: yes
    code_folding: hide
---

# Abstract
Eight samples in this run to be technical replicates (from the back of the eye). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      comment = "")
```

```{r}
library(biomaRt)
library(Seurat)
library(scCustomize)
library(sctransform)
library(glmGamPoi)
library(tidyverse)
library(corrplot)
library(tidyverse)
library(reshape2)
library(future)
options(future.globals.maxSize = 800 * 1024^2) # tested, use 600M
library(openxlsx)
library(patchwork)
library(HGNChelper)
library(httpgd)
Sys.setenv(PROJ_LIB = "/home/ubuntu/anaconda3/envs/r4.2.1/share/proj")
Sys.setenv(GDAL_DATA = "/home/ubuntu/anaconda3/envs/r4.2.1/share/gdal")
source("/mnt/data/apps/sc-type/R/gene_sets_prepare.R")
source("/mnt/data/apps/sc-type/R/sctype_score_.R")
setwd('/home/ubuntu/Ocular_snRNAseq_Aflibercept_NHP_1')
set.seed(123)
```


# Load and combine sample feature_bc_matrix from cellranger 
```{r, eval = FALSE}
rds_path <- list.files('./nf-core/output/cellranger/cellranger/mtx_conversions/',
    pattern = "GTSAMPLE0862-[0-9]_matrix\\.rds", full.names = TRUE)
rds_names <-  gsub("_matrix.rds$", "", basename(rds_path))
rds_list <- lapply(rds_path, readRDS)
names(rds_list) <- rds_names

# combine samples
seuratObj <- merge(rds_list[[1]], y = rds_list[2:8], add.cell.ids = names(rds_list), project = "ocular_Aflibercept")
# add sample names
cells <- rownames(seuratObj@meta.data)

seuratObj$sample <- sub("_.*", "", cells)
seuratObj$tissue <- "posterior"
```

# QC and selecting cells for further analysis

## Add %mitochondrial counts and cell cycle scores 
```{r, eval = FALSE}
## mitochondrial genes:
mt_genes <- c("ND1", "ND2", "COX1", "COX2", "ATP8", "ATP6", "COX3", "ND3", "ND4L", "ND4", "ND5", "ND6", "CYTB")

seuratObj[["percent.mt"]] <- PercentageFeatureSet(seuratObj, features = mt_genes)

s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

seuratObj <- CellCycleScoring(seuratObj, s.features = s.genes, g2m.features = g2m.genes)
```

## QC
- Filter cells by:  
1, umi/nCount_RNA, low-quality cells (very low UMI counts) or doublets (very high UMI counts)  
2, gene/nFeature_RNA, remove potential low-quality cells and multiplets as above  
3, percent.mt, note that this step directly removes cells with high mt, the regress.out function later   
removes variations from mt and cell-cyle gene expression.   
4, our cutoff is chosen by visulizing data in Loupe browser    
5
== Load filtered Seurat obj instead of re-run the following steps 
```{r}
seuratObj <- readRDS("ocular_obj.rds")
```

- The object after pre-process
```{r}
seuratObj
table(seuratObj[[]]$sample)
head(seuratObj, 5)
```

### QC stats
```{r}
summary(seuratObj$nCount_RNA)
summary(seuratObj$nFeature_RNA)
summary(seuratObj$percent.mt)
```

UMI range is (`r min(seuratObj$nCount_RNA)`, `r max(seuratObj$nCount_RNA)`)
nfeatue range is (`r min(seuratObj$nFeature_RNA)`, `r max(seuratObj$nFeature_RNA)`)

Retinal ganglion cells are larger, results in a greater amount of total RNA per cell. 
Skip the nCount_RNA and nFeature_RNA filter as the gRed team did in their 2023 Cell Genomics paper.

Only remove cells with >5% mitochondrial expressions: `r length(seuratObj$percent.mt[seuratObj$percent.mt > 5])`

```{r, eval = FALSE}
seuratObj <- subset(seuratObj, subset = percent.mt < 5)
```

### QC metrics as a violin plot 
- by initial clusters
```{r, fig.height=10, fig.width=12}
VlnPlot(seuratObj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), pt.size = 0, ncol = 1)
```

- by samples
```{r, fig.height=6, fig.width=12}
VlnPlot(seuratObj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
    group.by = 'sample',
    pt.size = 0.2, ncol = 3)
```


## SCTransform normalization, PCA and Clustering 
```{r, eval = FALSE}
# regressing out unwanted variation (mitochondrial, cell cyle, etc.)
seuratObj <- SCTransform(seuratObj, method = "glmGamPoi", vst.flavor = "v2",
        vars.to.regress = c("percent.mt", "S.Score", "G2M.Score"), verbose = TRUE)

seuratObj <- RunPCA(seuratObj, npcs = 40)
# ElbowPlot(seuratObj, ndims = 40)
seuratObj <- RunUMAP(seuratObj, reduction = "pca", dims = 1:28)
seuratObj <- FindNeighbors(seuratObj, reduction = "pca", dims = 1:28)
seuratObj <- FindClusters(seuratObj, resolution = 0.5)
```

== Save processed object
```{r, eval = FALSE}
saveRDS(seuratObj, file = "ocular_obj.rds")
```

## Visualize initial clusters
```{r, fig.height=7, fig.width=14}
p1 <- DimPlot(seuratObj, reduction = "umap", group.by = "sample") + theme(legend.position = "top")
p2 <- DimPlot(seuratObj, reduction = "umap", label = TRUE, repel = TRUE) + theme(legend.position = "top")
p1 + p2
```

```{r, fig.height=10, fig.width=14}
DimPlot(seuratObj, reduction = "umap", label = FALSE, split.by = "sample", ncol = 4) + theme(legend.position = "top")
```

## Number of cells in each cluster
```{r}
plot_clusters <- function(srat) {
    count_table <- table(srat@meta.data$seurat_clusters, srat@meta.data$sample)
    count_mtx <- as.data.frame.matrix(count_table)
    count_mtx$clusters <- rownames(count_mtx)
    melt_mtx <- melt(count_mtx)

    celltype_size <- aggregate(value ~ clusters, data = melt_mtx, FUN = sum)
    celltype_size <- celltype_size %>% dplyr::arrange(value)
    celltype_size$clusters <- factor(celltype_size$clusters, levels = celltype_size$clusters)
    melt_mtx$clusters <- factor(melt_mtx$clusters, levels = celltype_size$clusters)
    colnames(melt_mtx)[2] <- "sample"

    p1 <- ggplot(celltype_size, aes(y = clusters, x = value)) +
    geom_bar(position = "dodge", stat = "identity", fill = "grey60") +
    theme_bw() + scale_x_log10() + xlab("Cells per cluster, log10 scale") + ylab("")

    p2 <- ggplot(melt_mtx, aes(x = clusters, y = value, fill = sample)) +
        geom_bar(position = "fill", stat = "identity") + theme_bw() + coord_flip() +
        scale_fill_brewer(palette = "Set2") +
        ylab("Fraction of cells in each sample") + xlab("cluster number") + theme(legend.position = "top")

    p2 + p1 + plot_layout(widths = c(3, 1))
}
```
```{r fig.height=8, fig.width=14}
plot_clusters(seuratObj)
```


# Cell type annotation

## Apply marker based tool `sc-type`
```{r}
# prepare marker genes from the database, category: Immune system, Liver, Pancreas, Kidney, Eye, Brain
# gs_list <- gene_sets_prepare("/mnt/data/apps/sc-type/ScTypeDB_short.xlsx", "Eye")

# use a customized marker list
gs_list <- gene_sets_prepare("./ocular_customized_markers.xlsx", "Eye")
# get cell-type by cell matrix
# NOTE: scRNAseqData parameter should correspond to your input scRNA-seq matrix.
# In case Seurat is used, it is either pbmc[["RNA"]]@scale.data (default),
# pbmc[["SCT"]]@scale.data, in case sctransform is used for normalization,
# or pbmc[["integrated"]]@scale.data, in case a joint analysis of multiple single-cell datasets is performed.
es.max <- sctype_score(scRNAseqData = seuratObj[['SCT']]@scale.data, scaled = TRUE,
                      gs = gs_list$gs_positive, gs2 = gs_list$gs_negative)

# merge by cluster
metadata <- seuratObj@meta.data
clusters <- unique(seuratObj@meta.data$seurat_clusters)
cL_resutls <- do.call("rbind",
    lapply(
        clusters,
        function(cl) {
            es.max.cl <- sort(rowSums(es.max[, rownames(metadata[metadata$seurat_clusters == cl, ])]),
                decreasing = TRUE)
            cell_scores <- data.frame(cluster = cl,
                    type = names(es.max.cl),
                    scores = es.max.cl,
                    ncells = sum(metadata$seurat_clusters == cl))
            head(cell_scores, 10)
        }
    )
)
sctype_scores <- cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)

# set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] <- "Unknown"
data.frame(sctype_scores)

# add cell type annotations to obj
seuratObj@meta.data$celltypes <- ""
for (j in unique(sctype_scores$cluster)){
  cl_type <- sctype_scores[sctype_scores$cluster == j, ]
  seuratObj@meta.data$celltypes[seuratObj@meta.data$seurat_clusters == j] <-
    as.character(cl_type$type[1])
}
```


## Show cell types on UMAP
```{r, fig.height=14, fig.width=14}
DimPlot(seuratObj, reduction = "umap", label = TRUE, repel = TRUE,
    group.by = 'celltypes') +
    theme(legend.position = "top")
```

```{r, fig.height=14, fig.width=14}
DimPlot(seuratObj, reduction = "umap", label = TRUE, repel = TRUE, split.by = 'sample', ncol = 4,
    group.by = 'celltypes') +
    theme(legend.position = "top")
```

## Number of cells in each cell type
```{r}
plot_celltypes <- function(srat) {
    count_table <- table(srat@meta.data$celltypes, srat@meta.data$sample)
    count_mtx <- as.data.frame.matrix(count_table)
    count_mtx$celltypes <- rownames(count_mtx)
    melt_mtx <- melt(count_mtx)

    celltype_size <- aggregate(value ~ celltypes, data = melt_mtx, FUN = sum)
    celltype_size <- celltype_size %>% dplyr::arrange(value)
    celltype_size$celltypes <- factor(celltype_size$celltypes, levels = celltype_size$celltypes)
    melt_mtx$celltypes <- factor(melt_mtx$celltypes, levels = celltype_size$celltypes)
    colnames(melt_mtx)[2] <- "sample"

    p1 <- ggplot(celltype_size, aes(y = celltypes, x = value)) +
    geom_bar(position = "dodge", stat = "identity", fill = "grey60") +
    theme_bw() + scale_x_log10() + xlab("Cells per celltype, log10 scale") + ylab("")

    p2 <- ggplot(melt_mtx, aes(x = celltypes, y = value, fill = sample)) +
        geom_bar(position = "fill", stat = "identity") + theme_bw() + coord_flip() +
        scale_fill_brewer(palette = "Set2") +
        ylab("Fraction of cells in each sample") + xlab("celltype number") + theme(legend.position = "top")

    p2 + p1 + plot_layout(widths = c(3, 1))
}
```

```{r fig.height=6, fig.width=14}
plot_celltypes(seuratObj)
```



# Cluster markers analysis
== Load all cluster markers obj
```{r}
all.markers <- readRDS("all.cluster_markers.rds")
head(all.markers)
```

## Find all markers
```{r, eval = FALSE}
seuratObj <- PrepSCTFindMarkers(seuratObj, assay = "SCT", verbose = TRUE)
plan("multisession", workers = 4)
all.markers <- FindAllMarkers(seuratObj, assay = "SCT", verbose = TRUE)
plan(sequential)
```

## Save marker genes
```{r, eval = FALSE}
all.markers <- all.markers %>% relocate(gene)
write_tsv(all.markers, "all.cluster_markers.tsv")
top20 <- all.markers %>%
    group_by(cluster) %>%
    top_n(n = 25, wt = avg_log2FC)
write_tsv(top20, "top25.cluster_markers.tsv")
saveRDS(all.markers, file = "all.cluster_markers.rds")
```

## Heatmap with top genes {.tabset}

### Markers for each Cluster
- showing expression in 10% random cells for each cluster  
```{r, fig.height=20, fig.width=12}
Idents(seuratObj) <- "seurat_clusters"

top10 <- all.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC)

DoHeatmap(subset(seuratObj, downsample = 100), features = top10$gene, size = 3) +
    theme(legend.position = 'none', plot.margin = margin(t = 20))
```

# re-annotate cells by virtural inspection on markers

## RETINAL GANGLION CELLS (RGCs) ===============================================

```{r}
sn <- seuratObj
DefaultAssay(sn) <- "RNA"
table(Idents(sn))
# hgd()
# hgd_view()
DimPlot(sn, reduction = "umap", label = TRUE, repel = TRUE) + theme(legend.position = "top")
# p <- FeaturePlot(sn, split.by = "sample", combine = F, features = c("ENSMFAG00000036911"))
# cowplot::plot_grid(plotlist = p, ncol=2)

# check transgene Aflibercept first
FeaturePlot(sn, features = c("Aflibercept"))
VlnPlot(sn, features = "Aflibercept")
    # + scale_y_continuous(limits = c(0,30))

# RGCs are the projection neurons of the retina
# markers of RGCs are:

# RBPMS, which is not correctly annotated in macaque reference genome 
# transcript ID for this is "ENSMFAG00000036911"
FeaturePlot(sn, features = c("ENSMFAG00000036911"))
VlnPlot(sn, features = "ENSMFAG00000036911")

# glutamate transporter VGLUT2 (SLC17A6 gene)
VlnPlot(sn, features = "SLC17A6")

# gamma synuclein (SNCG)
VlnPlot(sn, features = "SNCG")

# synaptotagmin 2 (SYT2)
VlnPlot(sn, features = "SYT2")

# correlate percentage of cells expressing markers
perc_exp <- Percent_Expressing(seurat_object = seuratObj, features = c("Aflibercept", "ENSMFAG00000036911", "SNCG", "SYT2"))
# Calculate the correlation matrix
cor_matrix <- cor(t(perc_exp), method = "pearson")

# Create a matrix for p-values
p_values <- cor.mtest(t(perc_exp), method = "pearson")$p

# Create a correlation plot
corrplot(cor_matrix, type = "upper", method = "circle", tl.col = "black", addCoef.col = "black", cl.cex = 1.5,number.cex=1.5,
    p.mat = p_values, sig.level = c(.001, .01, .05), pch.cex = .9, insig = "label_sig", pch.col = "white")
```